---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Below is the R code which loads and preprocess the data.

```{r fig.path='figure/'}

df.data <-read.csv("activity.csv")

# create a Datetime class Date column 
df.data["Date"] <- NA
df.data$Date <- as.POSIXct(as.character(df.data$date), format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

The below R code calculates the total number of steps taken per day. Missing values are removed.

```{r}

# sum steps by Date and name the sum column 'TotalSteps' 
df.totalSteps.day.rmNA <-with(df.data, aggregate(steps, by=list(Date = Date), FUN=sum, na.rm=TRUE))

names(df.totalSteps.day.rmNA)[2] <- "TotalSteps"


```

The below R code produces a histogram that illustrates the total number of steps taken per day.

```{r fig.path='figure/'}
library("ggplot2")

# produce the histogram
ggplot(data=df.totalSteps.day.rmNA) + 
    geom_histogram(aes(x=Date, y=TotalSteps), fill="blue", stat="identity")+
    xlab("Date") +
    ylab("Total Steps") +
    ggtitle(expression("Total Steps per Day")) 

```

Mean total steps per day:

```{r}
mean(df.totalSteps.day.rmNA$TotalSteps)


```

Median total steps per day:

```{r}
median(df.totalSteps.day.rmNA$TotalSteps)
```
## What is the average daily activity pattern?

The below R code creates a line graph which illustrates the daily activity pattern (mean steps by 5-minute interval). Missing values are removed.

```{r fig.path='figure/'}

# average the steps by interval and name the average column as 'MeanSteps'
df.meanSteps.interval.rmNA <-with(df.data, aggregate(steps, by=list(Interval = interval), 
                                                     FUN=mean, na.rm=TRUE))

names(df.meanSteps.interval.rmNA)[2] <- "MeanSteps"


# create a interval column of Datetime class

df.meanSteps.interval.rmNA["TimeInterval"] <- NA


df.meanSteps.interval.rmNA$TimeInterval <- strptime(
    sprintf("%04d", df.meanSteps.interval.rmNA$Interval),"%H%M")

# produce the line graph

with(df.meanSteps.interval.rmNA, plot(TimeInterval, MeanSteps, type="l", ylab=expression("Average steps across all days"),xlab="Interval", main = expression("Average Steps by Interval")))

```

Highest number of mean steps is observed at this 5min interval:

```{r}
df.meanSteps.interval.rmNA[which(df.meanSteps.interval.rmNA[,2] == max(df.meanSteps.interval.rmNA$MeanSteps)),1]
```


## Imputing missing values

Below R code calculates the total number of rows with NA values in the dataset

```{r}
sum(is.na(df.data$steps))# shows count of NA

```


The below R code implements the strategy of filling missing values with mean number of steps for the interval.

```{r fig.path='figure/'}

# merge in the mean steps per interval from previous section, identify position of NAs and replace NAs with means steps
df.data.naFilled <- merge(df.data, df.meanSteps.interval.rmNA, by.x = "interval", by.y="Interval" )

v_na.Steps <- which(is.na(df.data.naFilled$steps)== TRUE)

df.data.naFilled[v_na.Steps,2] <- df.data.naFilled[v_na.Steps,5]


# sum up steps by Date and name the sum column 'TotalSteps'
df.data.naFilled.sum <-with(df.data.naFilled, aggregate(steps, by=list(Date = Date), 
                                                 FUN=sum))

names(df.data.naFilled.sum)[2] <- "TotalSteps"

# produce the histogram
ggplot(df.data.naFilled.sum) + 
    geom_histogram(aes(x=Date, y=TotalSteps), fill="green", stat="identity")+
    xlab("Date") +
    ylab("Total Steps") +
    ggtitle(expression("Total Steps per Day")) 

```

Mean total steps after imputing missing values
```{r}
mean(df.data.naFilled.sum$TotalSteps)
```


Median total steps after imputing missing values
```{r}
median(df.data.naFilled.sum$TotalSteps)
```


After imputing missing values, the mean and median of total steps increased from previous estimates. The mean and median of total steps converged to the same value.

## Are there differences in activity patterns between weekdays and weekends?

Below is the R code which creates the 'weekday' or 'weekend' factor column  (Cat) and the line graphs which illustrate the difference in mean steps taken between weekday and weekend.

```{r fig.path='figure/'}

# create a new factor column Cat for weekend or weekday factors
df.data.naFilled["Cat"] <- NA

df.data.naFilled$Cat <- ifelse (weekdays(df.data.naFilled$Date) %in% c("Saturday", "Sunday"), 
                                "weekend", "weekday")

df.data.naFilled$Cat <- as.factor(df.data.naFilled$Cat)

# average steps by interval and name the column 'MeanSteps'
df.data.naFilled.mean <-with(df.data.naFilled, aggregate(steps, by=list(TimeInterval = interval, Cat = Cat), 
                                                        FUN=mean))

names(df.data.naFilled.mean)[3] <- "MeanSteps"

df.data.naFilled.mean$TimeInterval <- strptime(sprintf("%04d", df.data.naFilled.mean$TimeInterval), "%H%M")

# load scales library to manipulate x scales
library("scales")

# plot the 2 line graphs
ggplot(df.data.naFilled.mean, aes(x=TimeInterval, y=MeanSteps)) +
    geom_line() +
    xlab("Interval") +
    ylab("Average Steps") +
    ggtitle(expression("Average Step by Intervals across Weekdays and Weekend")) + 
    scale_x_datetime(labels = date_format("%H:%M"),breaks = "2 hour") +
    facet_grid(Cat~ .)

```

From the plot above, in general, it can be observed that less steps are taken over weekend. 